Safe Transferable Regions

### 2017-01-14 paper/related.tex
Edits to appendix. CR04 added as related work.

Tofte and Talpin's approach~\cite{tofte97} uses compiler-managed lexically
scoped (stack) regions (as a replacement for GC).
Our type inference is analogous to their's in some respects,
while differing in others.
Their inference algorithm only generates equality constraints, solvable via unification.
Our type inference algorithm generates partial order outlives constraints.
% which are required to capture relationships between lifetimes of transferable regions and stack regions.
Consequently, our constraint solving algorithm is
more sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

@@

% Although there is some support for local
% type inference, region types for methods and classes need to be
% written explicitly. In contrast, our region type inference that
% eliminates any such need.

~\cite{HMN01} proposes a flow-sensitive approach for first-order programs
% to overcome some of the drawbacks of~\cite{tofte97}
to generalize~\cite{tofte97}'s approach to dynamic regions.
However, dynamic regions are still not first-class values of the language,
and reference counting is needed to ensure memory safety.
%
\cite{gpu14} uses regions to safely transfer data between the CPU and GPU
in the context of Scheme. However, their setting only includes lexically-scoped
regions for which Tofte and Talpin-style analysis~\cite{tofte97}
suffices. In contrast, we provide first-class support for
transferable regions with dynamic lifetimes.
% We require this generality in order to support streaming query operators, such as the
% one shown in Fig.~\ref{fig:motivating-eg}.

>>>>

Tofte and Talpin's approach~\cite{tofte97} uses compiler-managed lexically
scoped (stack) regions (as a replacement for GC).
Our type inference is analogous to theirs in some respects,
while differing in others.
Their inference algorithm only generates equality constraints, solvable via unification.
Our type inference algorithm generates partial order outlives constraints.
% which are required to capture relationships between lifetimes of transferable regions and stack regions.
Consequently, our constraint solving algorithm is
more sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

@@

~\cite{HMN01} proposes a flow-sensitive approach for first-order
programs to generalize~\cite{tofte97}'s approach to dynamic
regions.~\cite{CR04} describes a flow-insensitive and
context-sensitive analysis that transforms Java programs to use
(dynamic) regions.  However, neither~\cite{HMN01} nor~\cite{CR04}
supports dynamic regions as first-class objects; they cannot be stored
in data structures or passed to methods. Furthermore,
while~\cite{HMN01} requires reference counting to ensure memory
safety,~\cite{CR04} comes with no formal safety guarantees.
%
\cite{gpu14} uses regions to safely transfer data between the CPU and GPU
in the context of Scheme. However, their setting only includes lexically-scoped
regions for which Tofte and Talpin-style analysis~\cite{tofte97}
suffices. In contrast, we provide first-class support for
transferable regions with dynamic lifetimes.
% We require this generality in order to support streaming query operators, such as the
% one shown in Fig.~\ref{fig:motivating-eg}.

### 2017-01-14 paper/type-system.tex
Edits to appendix. CR04 added as related work.

Note that classes in \FB are independently parameterized over types
and regions. While this design decision has a downside that it allows
type variables to only denote the types of objects contained in a
single region, it yeilds benefits that far outweigh the costs. In
particular, it lets us support region-polymorphic higher-order
functions as class fields.  This allows, for example, a generic class,
whose type parameters are $a$ and $b$, to contain a region-polymorphic
function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
to the point that inference becomes practical
(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
type systems suggests, type inference is pivotal to the practical
utility of sophisticated type systems.

@@

The type rules distinguish between the \C{new} expressions that create
objects of  the \RgnZ class, and \C{new} expressions that create
objects of other classes. The rule for the latter relies on an
auxiliary definition called $\fields$ (undefined for $\RgnZ$ class)
that returns the sequence of type bindings for fields (instance
variables) of a given class type.  Like in FGJ, the names and types of
a constructor's arguments in \FB are same as the names and types of
its class's fields.

@@

The type rule for method application uses auxiliary definitions
$\mtype$ and $\bound$ which originate from the FGJ calculus - the
former gives the type of a class method and the latter defines the
bound of a type variable. Method application involves instantiation of
method's region parameters, and the type rule requires the first
region parameter to be instantiated with the current allocation
context ($r$). The type rule for function application does similarly.
This requirement ensures the safety of closures as demonstrated by the
following example.

@@

The method well-formedness rule makes use of an auxiliary definition
$\override$ that judges whether the current method is a valid
overriding of any eponymous method from the super class. The type
environment is extended with a binding that binds the $\thisZ$ keyword
to the type of the current class. Note that the the type of the
current method as accessed via $\thisZ$ (i.e., $\thisZ.m$) is
region-parametric, thus admitting region-polymorphic recursion (i.e.,
recursive call to $m$ can have different region arguments than $m$).

@@

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
additional langauge constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
while allocating a (static) region, and \C{open} reduces to \C{opened}
while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.

Operational semantics of \FB defines a four-place small-step reduction
relation of the form shown below:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
\end{smathpar}
The typestate of regions is tracked by a finite map ($\mem$) from
locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
relation relates an expression ($e$) and a typestate map ($\mem$) to a
reduced expression ($e'$) and an updated typestate map ($\mem'$). The
semantics gets ``stuck'' if $e$ attempts to access an object whose
allocation region is not present in $\rhoenv$, or if $e$ tries to
\C{open} a transferable region that is not mapped to an appropriate
typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
operation on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$). An illustrative subset of operational semantics
that formalize the intuitions desribed above is shown in
Fig.~\ref{fig:fb-opsem}.  Fig.~\ref{fig:fb-opsem-1} of the appendix
contains rest of the rules.

>>>>

Note that classes in \FB are independently parameterized over types
and regions. While this design decision has a downside that it allows
type variables to only denote the types of objects contained in a
single region, it yields benefits that far outweigh the costs. In
particular, it lets us support region-polymorphic higher-order
functions as class fields.  This allows, for example, a generic class,
whose type parameters are $a$ and $b$, to contain a region-polymorphic
function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
to the point that inference becomes practical
(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
type systems suggests, type inference is pivotal to the practical
utility of sophisticated type systems.

@@

The type rules distinguish between the \C{new} expressions that create
objects of  the \RgnZ class, and \C{new} expressions that create
objects of other classes. The rule for the latter relies on an
auxiliary definition\footnote{All auxiliary definitions we use in this
exposition originate from the FGJ calculus.} called $\fields$
(undefined for $\RgnZ$ class) that returns the sequence of type
bindings for fields (instance variables) of a given class type.  Like
in FGJ, the names and types of a constructor's arguments in \FB are
same as the names and types of its class's fields. The type rule for
the field access expression ($e.f_i$) also uses $\fields$ and another
definition called $\bound$, which returns the bound of a type variable
($\bound$ is an identity function for concrete types).

@@

The type rule for method application uses an auxiliary definition
$\mtype$ that gives the type of a class method. Method application
involves instantiation of method's region parameters, and the type
rule requires the first region parameter to be instantiated with the
current allocation context ($r$). The type rule for function
application does similarly.  This requirement ensures the safety of
closures as demonstrated by the following example.

@@

The method well-formedness rule makes use of an auxiliary definition
$\override$ that judges whether the current method is a valid
overriding of any eponymous method from the super class. The type
environment is extended with a binding that binds the $\thisZ$ keyword
to the type of the current class. Note that the the type of the
current method as accessed via $\thisZ$ (i.e., $\thisZ.\C{m}$) is
region-parametric, thus admitting region-polymorphic recursion (i.e.,
recursive call to $\C{m}$ can have different region arguments than
$\C{m}$).

@@

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
additional language constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
while allocating a (static) region, and \C{open} reduces to \C{opened}
while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.

Operational semantics of \FB defines a four-place small-step reduction
relation of the form shown below:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
\end{smathpar}
The typestate of regions is tracked by a finite map ($\mem$) from
locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
relation relates an expression ($e$) and a typestate map ($\mem$) to a
reduced expression ($e'$) and an updated typestate map ($\mem'$). The
semantics gets ``stuck'' if $e$ attempts to access an object whose
allocation region is not present in $\rhoenv$, or if $e$ tries to
\C{open} a transferable region that is not mapped to an appropriate
typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
operation on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$). An illustrative subset of operational semantics
that formalize the intuitions described above is shown in
Fig.~\ref{fig:fb-opsem}.  Fig.~\ref{fig:fb-opsem-1} of the appendix
contains rest of the rules.

### 2017-01-14 paper/typesystem-proofs.tex
Edits to appendix. CR04 added as related work.

Consistency between $\Delta$ and $\mem$ is preserved by the reduction
relation:

\begin{lemma}[\textbf{consistency preservation}]
\label{lem:consistency}
$\forall(e,\Delta,\mem,\phicx,\rgn,\tau)$. if $\consistent(\Delta,\mem)$
and $\tywf{\Delta}{\phicx}$ %and $\rgn \in \Delta$ 
and $\hastyp{(\Delta,\cdot,\phicx),\cdot,\rgn}{e}{\tau}$ 
and $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then
$\consistent(\Delta,\mem')$.
\end{lemma}
\begin{proof}
Proof is by induction on $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$.
For the rules where $\mem'=\mem$, proof is trivial. For the rules
where $\mem'$ is a result of executing a subexpression under
$\Delta$, proof follows from the inductive hypothesis. Remaining rules
are discussed below:
% \begin{smathpar}
% \begin{array}{cl}
%   \forall(\Delta,\mem,\phicx,\tau).~\consistent(\Delta,\mem)
%   ~\conj ~\tywf{\Delta}{\phicx}
%   ~\conj ~\hastyp{(\Delta,\cdot,\phicx),\cdot}{e_0}{\tau} & \\
%   \hspace*{1in}
%   ~\redstoo{\Delta}{(e_0,\mem)}{(e_0',\mem')}
%   ~\Rightarrow~ \consistent(\Delta,\mem') & IH\\
% \end{array}
% \end{smathpar}
\begin{itemize}
  \item Rule $\rulelabel{LetRegionBegin}$: All live regions in $\mem$
  are also live in $\mem'$. Proof follows
  \item Rule $\rulelabel{LetRegion}$: By inductive hypothesis, $\mem'$
  is consistent with $\Delta$. Hence, it is consistent
% is consistent with $\Delta \cup \{\loc\}$. Hence, it is consistent
% with $\Delta$.
  \item Rule $\rulelabel{LetRegionEnd}$: Here, $\loc$ is set to
  $\XFERRED$ in $\mem'$. However, since $e$ is the \C{letd}
  expression, and the type rule for the \C{letd} expression gives us
  $\loc \not\in \Delta$. Hence $\Delta$ is consistent with $\mem'$.
  \item Rule $\rulelabel{NewRegion}$: $\mem'$ extends $\mem$ with a
  new binding. Consistency is trivially preserved.
% \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\mem(\loc)=\USED$
% and $\consistent(\Delta,\mem)$, $\loc \not\in \Delta$. The proof
% follows from inductive hypothesis.
  \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\consistent(\Delta,
  \mem)$, we have $\consistent(\Delta \cup \{\loc\}, \mem[\loc\mapsto 
  \LIVE])$. By inductive hypothesis, $\consistent(\Delta \cup \{\loc\}, 
  \mem')$. Hence, $\consistent(\Delta, \mem')$.
  \item Rule $\rulelabel{Open}$: As $\mem'=\mem[\loc\mapsto\LIVE]$,
  hence $\Delta$ remains consistent with $\mem'$. Here, we also take
  note of the fact that if the result expression (\C{opened}) is
  tagged with a typestate of $\USED$, then $\loc\notin \Delta$.
  \item Rule $\rulelabel{Opened}$: Inductive hypothesis guarantees
  that $\Delta$ is consistent with $\mem'$.
% that $\Delta\cup\{\loc\}$ is consistent with $\mem'$. Hence,
% $\Delta$ is consistent with $\mem'$. 
  We also take note of the fact that the typestate tagged with the
  $\C{opened}$ expression remains invariant during reduction.
  \item Rule $\rulelabel{OpenEnd}$: $\mem'=\mem[\loc\mapsto s]$, where
  $s$ is the typestate tagged with the \C{opened} expression. As clear
  from the $\rulelabel{Open}$ rule, $s$ can be either $\USED$ or
  $\LIVE$. The type rule for \C{opened} tells us that if $s$ is
  $\USED$ then $\loc\notin\Delta$. Hence, in either case $\Delta$
  remains consistent with $\mem$.
  \item Rules $\rulelabel{Transfer}$: changes binding for a non-live
  location $\loc$. Hence, consistency is preserved.
\end{itemize} 
\qed
\end{proof}

\begin{lemma}[value substitution preserves typing]
\label{lem:substitution}
$\forall(e,z,\tau_1,\tau_2,\Delta,\phicx)$, if $\hastyp{(\Delta,\cdot,
\phicx),\cdot[x\mapsto\tau_1]}{e}{\tau_2}$ and $\hastyp{(\Delta,\cdot,
\phicx),\cdot}{v}{\tau_1}$, then $\hastyp{(\Delta,\cdot,
\phicx),\cdot} {[v/x]e}{\tau_2}$.
\end{lemma}
\begin{proof}
The proof is by induction on typing derivation and follows on the
lines of similar proof for FGJ.
\qed
\end{proof}

@@

\end{itemize}
\qed
\end{proof}

@@

  \item \rulee{OpenEnd} case: The proof is very similar to the
  \C{letd} case.
  \qed
  \end{itemize}
\end{proof}

\begin{proof}[\textbf{Theorem~\ref{thm:fb-type-safety}}]
Follows from Lemmas~\ref{lem:consistency},~\ref{lem:progress},
and~\ref{lem:preservation}.
\qed
\end{proof}

>>>>

The rationale behind the $\consistent$ relation is to establish the
correspondence between the concrete memory ($\mem$) of the operational
semantics and its abstraction ($\Delta$) used by the static semantics.
We now prove that consistency between $\Delta$ and $\mem$ is preserved
by the reduction relation:

\begin{lemma}[\textbf{consistency preservation}]
\label{lem:consistency}
$\forall(e,\Delta,\mem,\phicx,\rgn,\tau)$. if $\consistent(\Delta,\mem)$
and $\tywf{\Delta}{\phicx}$ %and $\rgn \in \Delta$ 
and $\hastyp{(\Delta,\cdot,\phicx),\cdot,\rgn}{e}{\tau}$ 
and $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then
$\consistent(\Delta,\mem')$.
\end{lemma}
\begin{proof}
Proof is by induction on $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$.
For the rules where $\mem'=\mem$, proof is trivial. For the rules
where $\mem'$ is a result of executing a subexpression under
$\Delta$, proof follows from the inductive hypothesis. Remaining rules
are discussed below:
% \begin{smathpar}
% \begin{array}{cl}
%   \forall(\Delta,\mem,\phicx,\tau).~\consistent(\Delta,\mem)
%   ~\conj ~\tywf{\Delta}{\phicx}
%   ~\conj ~\hastyp{(\Delta,\cdot,\phicx),\cdot}{e_0}{\tau} & \\
%   \hspace*{1in}
%   ~\redstoo{\Delta}{(e_0,\mem)}{(e_0',\mem')}
%   ~\Rightarrow~ \consistent(\Delta,\mem') & IH\\
% \end{array}
% \end{smathpar}
\begin{itemize}
  \item Rule $\rulelabel{LetRegionBegin}$: All live regions in $\mem$
  are also live in $\mem'$. Proof follows
  \item Rule $\rulelabel{LetRegion}$: By inductive hypothesis, $\mem'$
  is consistent with $\Delta$. Hence, it is consistent
% is consistent with $\Delta \cup \{\loc\}$. Hence, it is consistent
% with $\Delta$.
  \item Rule $\rulelabel{LetRegionEnd}$: Here, $\loc$ is set to
  $\XFERRED$ in $\mem'$. However, since $e$ is the \C{letd}
  expression, and the type rule for the \C{letd} expression gives us
  $\loc \not\in \Delta$. Hence $\Delta$ is consistent with $\mem'$.
  \item Rule $\rulelabel{NewRegion}$: $\mem'$ extends $\mem$ with a
  new binding. Consistency is trivially preserved.
% \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\mem(\loc)=\USED$
% and $\consistent(\Delta,\mem)$, $\loc \not\in \Delta$. The proof
% follows from inductive hypothesis.
  \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\consistent(\Delta,
  \mem)$, we have $\consistent(\Delta \cup \{\loc\}, \mem[\loc\mapsto 
  \LIVE])$. By inductive hypothesis, $\consistent(\Delta \cup \{\loc\}, 
  \mem')$. Hence, $\consistent(\Delta, \mem')$.
  \item Rule $\rulelabel{Open}$: As $\mem'=\mem[\loc\mapsto\LIVE]$,
  hence $\Delta$ remains consistent with $\mem'$. Here, we also take
  note of the fact that if the result expression (\C{opened}) is
  tagged with a typestate of $\USED$, then $\loc\notin \Delta$.
  \item Rule $\rulelabel{Opened}$: Inductive hypothesis guarantees
  that $\Delta$ is consistent with $\mem'$.
% that $\Delta\cup\{\loc\}$ is consistent with $\mem'$. Hence,
% $\Delta$ is consistent with $\mem'$. 
  We also take note of the fact that the typestate tagged with the
  $\C{opened}$ expression remains invariant during reduction.
  \item Rule $\rulelabel{OpenEnd}$: $\mem'=\mem[\loc\mapsto s]$, where
  $s$ is the typestate tagged with the \C{opened} expression. As clear
  from the $\rulelabel{Open}$ rule, $s$ can be either $\USED$ or
  $\LIVE$. The type rule for \C{opened} tells us that if $s$ is
  $\USED$ then $\loc\notin\Delta$. Hence, in either case $\Delta$
  remains consistent with $\mem$.
  \item Rules $\rulelabel{Transfer}$: changes binding for a non-live
  location $\loc$. Hence, consistency is preserved.
\end{itemize} 
%\qed
\end{proof}

\begin{lemma}[value substitution preserves typing]
\label{lem:substitution}
$\forall(e,z,\tau_1,\tau_2,\Delta,\phicx)$, if $\hastyp{(\Delta,\cdot,
\phicx),\cdot[x\mapsto\tau_1]}{e}{\tau_2}$ and $\hastyp{(\Delta,\cdot,
\phicx),\cdot}{v}{\tau_1}$, then $\hastyp{(\Delta,\cdot,
\phicx),\cdot} {[v/x]e}{\tau_2}$.
\end{lemma}
\begin{proof}
The proof is by induction on typing derivation and follows on the
lines of similar proof for FGJ.
%\qed
\end{proof}

@@

\end{itemize}
%\qed
\end{proof}

@@

  \item \rulee{OpenEnd} case: The proof is very similar to the
  \C{letd} case.
  %\qed
  \end{itemize}
\end{proof}

\begin{proof}[\textbf{Theorem~\ref{thm:fb-type-safety}}]
Follows from Lemmas~\ref{lem:consistency},~\ref{lem:progress},
and~\ref{lem:preservation}.
%\qed
\end{proof}
